Effectful Semantics for Substructural SSA

### 2023-11-07 pldi24/paper.tex
Tap

\begin{itemize}
\item We have a syntax and type system for SSA-style programs. We include support for substructural types, and our type-theoretic syntax makes formulating inlining and rewriting easier than in the traditional CFG-based formulations. 
\item We show that our language has a compositional intepretation in any effectful category with an Elgot structure. We use this structure to give generic proofs of the expected substitution properties as well as prove the soundness of an equational theory for the language.
\item Our generic equational theory justifies a variety of powerful optimisation techniques, including E-graph rewriting, inlining, loop fusion, hoisting, and strength reduction. 
\item We show how Elgot monads give rise to models of our language. We use this to derive semantics from a general trace monad, and show that various monad transformers such as state transformers preserve the needed Elgot structure. 
\item We use these tools to demonstrate the existence of a variety of concrete models satisfying these categorical axioms. Starting first from simple languges such as state plus printing, we also show how more challenging semantics such as the TSO semantics for weak memory also fit into our framework. This thus gives rise to an SSA-based IR with support for weak memory operations, which is fully semantically-justified.
\item \TODO{We also have a small implementation, which \ldots}
\item Our weak memory semantics and our results about Elgot monads are formalised in the Lean theorem prover. \TODO{Other stuff}
\end{itemize}

@@

\TODO{define premonoidal category?}

@@

@@

\begin{definition}[Pomset] \
  \ \TODO{this}
\end{definition}

@@

@@

We'll start by considering the category \(\ms{HRel}\), which we will take to have morphisms of the form \(A \to \ms{Heap} \to \mc{P}^+(A \times \ms{Heap} \times \ms{Bytes}^* + \ms{Bytes}^{\leq \omega})\), which is the result of applying the state-transformer monad with state \(\ms{Heap}\) to the trace monad with the infinite monoid of lists of strings described in the previous section.

\TODO{Name for ND trace monad}

% \TODO{by the above, this is an \isotopessa model w/ nice properties; moving on!}

% We now define a separation logic with predicates having the following grammar:
% \begin{grammar}
%   <\(\varphi, \psi\)> ::= 
%     \(\varnothing\) 
%     \;|\; \(e \mapsto a\) 
%     \;|\; \(\varphi * \psi\)
%     \;|\; \(\forall x: A, \varphi\)
%     \;|\; \(\exists x: A, \varphi\)
%     \;|\; \(\varphi \land \psi\)
%     \;|\; \(\varphi \lor \psi\)
%     \;|\; \(\varphi \Rightarrow \psi\)
%     \;|\; \(\top\)
%     \;|\; \(\bot\)
% \end{grammar}
% where \(e, a\) denote regular \isotopessa expressions and \(A, B\) regular \isotopessa types.

% We can now do something somewhat tricky with \isotopessa dialects:
% \TODO{truly pure expressions \(\implies\) semantics for expressions}
% \TODO{giving us a semantics for separation logic stuff}
% \TODO{now we define the category \(\ms{Sep}\) w/ objects contexts + predicates}
% \TODO{tensor product, niceties \(\implies\) category, yey}
% \TODO{sliding}
% \TODO{therefore, substructural \textit{effectful} w/ nontrivial linear objects}

@@

@@

\TODO{start appendix}

>>>>

\begin{itemize}
\item We have a syntax and type system for SSA-style programs. We include support for substructural types, and our type-theoretic syntax makes formulating inlining and rewriting easier than in the traditional CFG-based formulations. 
\item We show that our language has a compositional intepretation in any effectful category with an Elgot structure. We use this structure to give generic proofs of the expected substitution properties as well as prove the soundness of an equational theory for the language.
\item Our generic equational theory justifies a variety of powerful optimisation techniques, including E-graph rewriting, inlining, loop fusion, hoisting, and strength reduction. 
\item We show how Elgot monads give rise to models of our language. We use this to derive semantics from a general trace monad, and show that various monad transformers such as state transformers preserve the needed Elgot structure. 
\item We use these tools to demonstrate the existence of a variety of concrete models satisfying these categorical axioms. Starting first from simple languges such as state plus printing, we also show how more challenging semantics such as the TSO semantics for weak memory also fit into our framework. This thus gives rise to an SSA-based IR with support for weak memory operations, which is fully semantically-justified.
\item \todo{We also have a small implementation, which \ldots}
\item Our weak memory semantics and our results about Elgot monads are formalised in the Lean theorem prover. \TODO{Other stuff}
\end{itemize}

@@

\begin{definition}[Binoidal Category] \
  \TODO{this}
\end{definition}

\begin{definition}[Premonoidal Category] \
  \TODO{this}
\end{definition}

@@

\TODO{find better notation than \(\Pi\)?}

@@

\begin{definition}[Pomset] 
  A \textbf{pomset} over a set of actions \(\mc{A}\) with a distinguished null action "\textbf{tick}" \(\delta \in \mc{A}\)
  \TODO{finish, clean}
\end{definition}

@@

Let \(\ms{Pom}\) denote the monoid of pomsets...

\TODO{stream action... trace monad \(\ms{Traces}\;\ms{Pom}\;\ms{Pom}\)...}

@@

We'll start by considering the category \(\ms{HRel}\), which we will take to have morphisms of the form \(A \to \ms{Heap} \to \mc{P}^+(A \times \ms{Heap} \times \ms{Bytes}^* + \ms{Bytes}^{\leq \omega})\), which is the result of applying the state-transformer monad with state \(\ms{Heap}\) to the trace monad with the monoid of lists of bytestrings described in the previous section.

\TODO{say something better than "bytestrings"}

\TODO{what is a \(\ms{Heap}\)}

\TODO{Use name \ms{Traces}}

\TODO{by the above, this is an \isotopessa model w/ nice properties; moving on!}

Given a set \(A\), we can define a predicate \(\varphi\) on \(A\), \(\mb{Heap}\) pairs to have the following type:
\TODO{this}
\TODO{Objects \((A, \varphi)\), \((B, \psi)\)}
\TODO{arrows}
\TODO{note on magic wand}
\TODO{Tensor product and separating conjunction}
\TODO{Coproduct, Elgot structure inherited from base category}
\TODO{sliding w/ properly restricted equality}

\TODO{so, it's a substructural effectful category mmkay}
\TODO{central = no printing}
\TODO{linear objects iff non duplicable}
\TODO{relevant morphisms iff deterministic}
\TODO{affine morphisms always (if no heap effects?)}

@@

\TODO{\cite{rvsdg}}

@@

\clearpage

\appendix

\section{Appendix}
